Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-16 PAPER.tex
revised conclusion based on aaron's edits

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop.png}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$ for fringe-rate filtered PAPER-64 data. Values are shown before (grey) and after (black) signal loss correction as a function of number of eigenmodes of $\hat{\textbf{C}}$ that are down-weighted. This regularization knob is tuned from $0$ modes (i.e. unweighted) to $21$ modes (i.e. full inverse covariance weighting). Over $\sim3$ orders of magnitude of signal loss results when using inverse covariance weighting. Also plotted for comparison are $2\sigma$ power spectrum upper limits for the unweighted case (dashed blue), inverse variance weighted case (red), and added-identity case (cyan). All three regularizations shown (black, red, cyan) perform similarly at their best.}
	\label{fig:sigloss_modeloop}
\end{figure*}

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure \ref{fig:ps1_data}. Again, black and grey points correspond to positive and negative power spectrum values respectively, with $2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim50$ in mK$^{2}$ (though we only use one baseline type in our analysis as opposed to $3$).

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Highest sensitivity power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and grey points correspond to positive and negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical noise level prediction based on observational parameters. This power spectrum result differs from \citet{ali_et_al2015} in that it only uses data from one type of baseline ($30$ m East/West baselines) instead of three. Additionally, an optimal fringe-rate filter was used instead of a degraded one, resulting in an increase in sensitivity but also an increase in signal loss. Major differences from previously published results stem from revisions regarding signal loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim50$ in mK$^{2}$.}
	\label{fig:ps1_data}
\end{figure*}

@@

Although current $21$ cm published power spectrum upper limits lie a $\sim$couple orders of magnitude above predicted EoR levels, ongoing analyses of deeper sensitivity datasets from PAPER, MWA, and LOFAR, as well as next generation instruments like HERA, expect to continue to push towards EoR sensitivities. As the field progresses towards a coveted detection, we have shown that it is crucial for future analyses to have a rigorous understanding of signal loss in an analysis pipeline, be able to accurately and robustly calculate both power spectrum and theoretical errors, and consistently undergo a comprehensive set of jackknive and null tests.

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances. Loss of the $21$ cm signal is especially significant if more eigenmodes are down-weighted than exist in the data (having an under-constrained covariance matrix). Hence, there exists a trade-off between sensitivity driven time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss. However, a signal-only simulation can underestimate loss by failing to account for spurious correlations between the data and signal. 
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are correlated. If the number of independent samples in a dataset is well-determined, bootstrapping is a simple and accurate way of estimating errors.
\item An EoR detection must pass all null tests. Similarly, performing jackknife tests along multiple axes of a dataset is necessary for confidence in a true EoR detection, and can also be used to tease out systematics.
\end{itemize}

As a consequence of our investigations, we have also revised power spectrum results from PAPER-64. For an analysis of only one baseline type, we quote a new $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim50$ in mK$^{2}$. The reasons for a previously underestimated limit and ways in which our new analysis differs can be summarized by the following:

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in \citet{ali_et_al2015}, was underestimated by a factor of $\sim$$1000$. Because PAPER-64 is limited by its number of samples, we use a regularized covariance weighting in our revised analysis to minimize loss, thereby increasing the original PAPER-64 limit in mK$^{2}$ by a factor of $\sim$$50$ instead of $1000$. Underestimated signal loss represents the bulk of our revision.
\item Power spectrum errors, originally computed by bootstrapping, were underestimated by a factor of $\sim$$2$ due to oversampling data with a reduced number of independent samples from fringe-rate filtering. They were also overestimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power spectra. Hence, the net change for our error estimation is negligible, though the intuition behind the previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding an increase in predicted sensitivity level by a factor of $\sim$$7$. We have verified our revised prediction extensively using pure noise simulations.
\end{itemize}

The $21$ cm future is exciting, as new experiments have sensitivities that expect to reach and surpass EoR levels, improved foreground mitigation/removal strategies are being developed, and simulations are being designed to better understand instruments. On the power spectrum analysis side, we recognize that there is still much work to do. Even with robust signal loss simulations and precise error calculations, we are currently limited by data volume and foreground suppression strategies. However, the future promises to bring higher sensitivity $21$ cm measurements, better foreground and instrumental models, more effective and less lossy weighting techniques --- and with all that, an improved understanding of reionization and our early Universe.

@@

\section{Acknowledgements}
The author would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA are supported by grants from the National Science Foundation (awards 0804508, 1129258, 1125558, 1440343, and 1636646). ARP, JCP, DCJ, and JEA would also like to acknowledge NSF support (awards 1352519, 1302774, 1401708, and 1455151, respectively). We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

>>>>

\subsubsection{Jackknife Tests}
\label{sec:JackknifeOverview}

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop.png}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$ for fringe-rate filtered PAPER-64 data. Values are shown before (grey) and after (black) signal loss correction as a function of number of eigenmodes of $\hat{\textbf{C}}$ that are down-weighted. This regularization knob is tuned from $0$ modes (i.e. unweighted) to $21$ modes (i.e. full inverse covariance weighting). Over $\sim3$ orders of magnitude of signal loss results when using inverse covariance weighting. Also plotted for comparison are $2\sigma$ power spectrum upper limits for the unweighted case (dashed blue), inverse variance weighted case (red), and added-identity case (cyan). All three regularizations shown (black, red, cyan) perform similarly at their best. Finally, a theoretical prediction for noise ($2\sigma$ error) is plotted as solid green.}
	\label{fig:sigloss_modeloop}
\end{figure*}

@@

Our revised, best to-date PAPER-64 power spectrum (using only one baseline separation type) is shown in Figure \ref{fig:ps1_data}. Again, black and grey points correspond to positive and negative power spectrum values respectively, with $2\sigma$ errors bars. Also plotted are the unweighted power spectrum upper limit (dashed blue) and theoretical prediction of noise (solid green). From this result, we quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim7$ in mK (though we only use one baseline type in our analysis as opposed to $3$). \cc{ref kolopanis limit instead later}

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_data_reg.png}
	\includegraphics[width=0.4\textwidth]{plots/ps2_data_reg.png}
	\caption{Power spectrum of PAPER-64 using $\hat{\textbf{C}}_{eff}$. Black and grey points correspond to positive and negative power spectrum values, respectively, with $2\sigma$ error bars also plotted. The dashed blue line is the unweighted power spectrum ($2\sigma$ upper limit). The solid green line is the theoretical noise level prediction based on observational parameters. This power spectrum result differs from \citet{ali_et_al2015} in that it only uses data from one type of baseline ($30$ m East/West baselines) instead of three. Major differences from previously published results stem from revisions regarding signal loss, bootstrapping, and the theoretical error computation. We quote a best $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim7$ in mK \cc{ref kolopanis limit instead later}.}
	\label{fig:ps1_data}
\end{figure*}

@@

Although current $21$ cm published power spectrum upper limits lie a couple orders of magnitude above predicted EoR levels, ongoing analyses of deeper sensitivity datasets from PAPER, MWA, and LOFAR, as well as next generation instruments like HERA, are expected to continue to push towards EoR sensitivities. As the field progresses towards a detection, we have shown that it is crucial for future analyses to have a rigorous understanding of signal loss in an analysis pipeline, be able to accurately and robustly calculate both power spectrum and theoretical errors, and consistently undergo a comprehensive set of jackknive and null tests.

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances (Section \ref{sec:SiglossOverview}). Loss of the $21$ cm signal is especially significant if more eigenmodes are down-weighted than exist in the data (having an under-constrained covariance matrix). Hence, there exists a trade-off between sensitivity driven time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss (Section \ref{sec:Sigloss}). However, a signal-only simulation (i.e. comparing an unweighted vs. weighted power spectrum of EoR) can underestimate loss by failing to account for spurious correlations between the data and signal.
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are significantly correlated (Section \ref{sec:ErrorOverview}). If the number of independent samples in a dataset is well-determined, bootstrapping is a simple and accurate way of estimating errors.
\item Meaningful null tests are vital to validate an EoR detection (Section \ref{sec:JackknifeOverview}). Similarly, performing jackknife tests along multiple axes of a dataset is necessary for confidence in an EoR detection, and can also be used to tease out systematics.
\end{itemize}

As a consequence of our investigations, we have also revised power spectrum results from PAPER-64. For an analysis of only one baseline type, we quote a new $2\sigma$ upper limit of $(158.0$ mK$)^{2}$ at $k=0.34$ $h$Mpc$^{-1}$, a higher limit than \citet{ali_et_al2015} by a factor of $\sim7$ in mK \cc{replace with kolopanis limit later}. The reasons for a previously underestimated limit and ways in which our new analysis differs can be summarized by the following:

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in \citet{ali_et_al2015}, was underestimated by a factor of $\sim$$1000$ for inverse covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section \ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a factor of $\sim7$ in mK higher than previous limits. Underestimated signal loss therefore represents the bulk of our revision.
\item Power spectrum errors, originally computed by bootstrapping, were underestimated by a factor of $\sim$$2$ due to oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. They were also overestimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a decrease in predicted sensitivity level by a factor of $\sim$$7$ (Section \ref{sec:PSSense}). We have verified our revised prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our power spectrum result, implying that the theoretical prediction for noise in \citet{ali_et_al2015} was too high for its data points.
\end{itemize}

The $21$ cm future is exciting, as new experiments have sensitivities that expect to reach and surpass EoR levels, improved foreground mitigation/removal strategies are being developed, and simulations are being designed to better understand instruments. On the power spectrum analysis side, robust signal loss simulations, precise error calculations, and comprehensive jackknife tests will play critical roles in accurate $21$ cm results. Looking ahead, the future promises to bring higher sensitivity measurements, better foreground and instrumental models, more effective and less lossy weighting techniques --- and with all that, an improved understanding of reionization and our early Universe.

@@

\section{Acknowledgements}
CC would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, JCP, DCJ, and JEA would also like to acknowledge NSF support (awards 1352519, 1302774, 1401708, and 1455151, respectively). We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}
